Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study

### 2021-03-01 submission/manuscript.rmd
proof reading

\newpage
\linenumbers
# Introduction
Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years [@wright2017impact]. It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs. By 2025, annual fractures and costs could rise by almost 50% [@Cole2008].

@@

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement [@Kent2019; @KentEnE2019]. Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates for the development of our prediction models.

For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1:4) subset of the risk subgroup subset. We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.

Residual study bias from unmeasured confounding can still be present in observational studies, which often may not be visible when evaluating propensity score adjustment. To account for that, we considered a set of 126 negative control outcome experiments, where a null effect was assumed to be true. We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest.

@@

<!-- Need to fix the numbers. See icpe abstract -->
A total of `r sum(incidenceOverall$maxTreatment) + sum(incidenceOverall$maxComparator)` patients were initially considered for our analyses. Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate (##TABLE1).

@@

@@

When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, Optum-DOD and Optum-EHR respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases. The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.

@@

\newpage
# Tables and figures
```{r, echo=FALSE}
incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
incidence <- incidenceOverall %>% 
  filter(analysisType == "matchOnPs_1_to_4") %>%
  mutate(
    database = case_when(
      database == "mdcr" ~ "MDCR",
      database == "optum_dod" ~ "Optum-DOD",
      database == "panther" ~ "Optum-EHR"
    ),
    treatmentDays = round(treatmentDays / 365),
    comparatorDays = round(comparatorDays / 365)
  ) %>%
  select(
    database,
    outcomeId,
    treatmentPersons,
    treatmentDays,
    treatmentOutcomes,
    comparatorPersons,
    comparatorDays,
    comparatorOutcomes
  ) %>%
  arrange(database, outcomeId) %>%
  mutate(
    database = ifelse(
      outcomeId == 101,
      database,
      ""
    ),
    outcomeId = case_when(
      outcomeId == 101 ~ "Hip fracture",
      outcomeId == 102 ~ "Major osteoporotic\nfracture",
      outcomeId == 103 ~ "Vertebral fracture"
    ),
  )

kable(
  incidence,
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Database",
    "Outcome",
    "Persons",
    "Person years",
    "Events",
    "Persons",
    "Person years",
    "Events"
  ),
  format.args = list(
    big.mar = ","
  )
) %>%
  add_header_above(c("", "", "Teriparatide" = 3, "Bisphosphonates" = 3))
  # kableExtra::add_indent(c(1, 4, 7))
```

>>>>

\newpage

\linenumbers
# Introduction
Osteoporosis is a common condition characterised by decreased bone density and associated with increased risk for fragility fractures, which affects almost 30% of women aged $\geq50$ years [@wright2017impact]. It was estimated that in 2005, over 2 million incident fractures occurred in the United States, with hip fractures accounting for more than 70% of the costs. By 2025, annual fractures and costs could rise by almost 50% [@Cole2008].

@@

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement [@Kent2019; @KentEnE2019]. Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates as for the development of the propensity score models.

For each outcome we used the derived prediction models to divide the population in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1:4) subset of the risk subgroup subset. We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.

Residual study bias from unmeasured confounding can still be present in observational studies, which often may not be visible when evaluating propensity score adjustment. To account for that, we considered a set of 126 negative control outcome experiments, where a null effect was assumed to be true. We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest [@Schuemie2013; @Schuemie2016].

@@

<!-- Need to fix the numbers. See icpe abstract -->
A total of `r sum(incidenceOverall$maxTreatment) + sum(incidenceOverall$maxComparator)` patients were initially considered for our analyses. Of those, `r sum(incidenceOverall$maxTreatment)` received teriparatide and `r sum(incidenceOverall$maxComparator)` received an oral bisphosphonate \@ref(tab:table1).

@@

hipRiskPerformance <- readRDS(here::here("data/raw/predictionPerformance.rds")) %>%
  filter(
    cohort == "EntirePopulation",
    stratOutcome == 101
  )

@@

The internally developed prediction models for hip fracture showed adequate performance achieving AUC of `r hipRiskPerformance %>% filter(database == "mdcr") %>% pull(auc) %>% d()`, `r hipRiskPerformance %>% filter(database == "optum_dod") %>% pull(auc) %>% d()` and `r hipRiskPerformance %>% filter(database == "panther") %>% pull(auc) %>% d()` in MDCR, Optum-DOD and Optum-EHR respectively. When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, Optum-DOD and Optum-EHR respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases. For the secondary outcomes (major osteoporotic fracture and vertebral fracture) we found no evidence of risk-based treatment effect heterogeneity. This was anticipated as overall effects of teriparatide on theses outcomes were not significant at the 95% confidence level. The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.

@@

\newpage
# Tables and figures
```{r table1, echo=FALSE, results='asis'}
incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
incidence <- incidenceOverall %>% 
  filter(analysisType == "matchOnPs_1_to_4") %>%
  mutate(
    database = case_when(
      database == "mdcr" ~ "MDCR",
      database == "optum_dod" ~ "Optum-DOD",
      database == "panther" ~ "Optum-EHR"
    ),
    treatmentDays = round(treatmentDays / 365),
    comparatorDays = round(comparatorDays / 365)
  ) %>%
  select(
    database,
    outcomeId,
    treatmentPersons,
    treatmentDays,
    treatmentOutcomes,
    comparatorPersons,
    comparatorDays,
    comparatorOutcomes
  ) %>%
  arrange(database, outcomeId) %>%
  mutate(
    database = ifelse(
      outcomeId == 101,
      database,
      ""
    ),
    outcomeId = case_when(
      outcomeId == 101 ~ "Hip fracture",
      outcomeId == 102 ~ "Major osteoporotic\nfracture",
      outcomeId == 103 ~ "Vertebral fracture"
    ),
  )

knitr::kable(
  incidence,
  format   = "latex",
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Database",
    "Outcome",
    "Persons",
    "Person years",
    "Events",
    "Persons",
    "Person years",
    "Events"
  ),
  format.args = list(
    big.mar = ","
  ),
   caption = "Number of patients, person years and outcomes for each outcome across all databases."
) %>%
  add_header_above(c("", "", "Teriparatide" = 3, "Bisphosphonates" = 3))
  # kableExtra::add_indent(c(1, 4, 7))
```
